---
title: "Week 11: Splines"
date: today
date-format: "DD/MM/YY"
format: pdf
execute: 
  warning: false
  message: false
---

# Overview

In this lab we'll be fitting a second-order P-Splines regression model to foster care entries by state in the US, projecting out to 2030. 

```{r}
library(tidyverse)
library(here)
library(rstan)
library(tidybayes)
source(here("code/getsplines.R"))
```

Here's the data

```{r}
d <- read_csv(here("data/fc_entries.csv"))
```

Let's make a plot highlighting trends over time by state. 


```{r}
d |> 
  ggplot(aes(year, ent_pc)) + 
  geom_line()+
  facet_wrap(~state)
```


## P-Splines 

Let's fit a hierarchical second-order P-Splines regression model to estimate the (logged) entries per capita over the period 2010-2017. The model we are fitting is

$$
\begin{aligned}
y_{st} &\sim N(\log \lambda_{st}, \sigma^2_{y,s})\\
\log \lambda_{st} &= \alpha_kB_k(t)\\
\Delta^2\alpha_k &\sim N(0, \sigma^2_{\alpha,s})\\
\log \sigma_{\alpha,s} &\sim N(\mu_{\sigma}, \tau^2)
\end{aligned}
$$
Where $y_{s,t}$ is the logged entries per capita for state $s$ in year $t$. We use cubic splines that have knots 2.5 years apart and are a constant shape at the boundaries. Standard normal priors are on standard deviations and hyperparameters. 

Get the data in the right format:

```{r}
years <- unique(d$year)
states <- unique(d$state)
N <- length(years)
S <- length(unique(d$state))
y <- d |> 
  select(state, year, ent_pc) |> 
  mutate(log_ent = log(ent_pc)) |> 
  select(-ent_pc) |> 
  pivot_wider(names_from = "state", values_from = "log_ent") |> 
  select(-year) |> 
  as.matrix()

pop <- d %>%
  dplyr::select(state, year, child_acs) %>%
  pivot_wider(names_from = "state", values_from = "child_acs") %>%
  dplyr::select(-year) %>% as.matrix()


Bs <- getsplines(years, I = 2.5)$B.ik

stan_data <- list(y = y,
                  B = Bs,
                  K = ncol(Bs),
                  N = N,
                  S = S)
```

Run the model

```{r}
mod <- stan(data = stan_data, file = "../code/models/fc_splines.stan")
```

## Future projection

Let's project forward entries per capita to 2030. 

```{r}
yhat <- mod %>%
  gather_draws(lambda[i,s]) %>%
  mutate(fit = exp(.value)) %>%
  median_qi() %>%
  mutate(year = years[i])

Bs_full <- getsplines(I = 2.5, 2000:2030)$B.ik

alphas <- extract(mod)[["alpha"]]
sigmas <- extract(mod)[["sigma"]]
sigma_ys <- extract(mod)[["sigma_y"]]

proj_years <- 2018:2030
K <- ncol(Bs)
K_full <- ncol(Bs_full)
proj_steps <- K_full - K

nsims <- nrow(alphas)

proj <- tibble()

alphas_proj <- array(NA, c(nsims, proj_steps, length(states)))

set.seed(1098)
# project the alphas
for(j in 1:length(states)){
  first_next_alpha <- rnorm(n = nsims,
                            mean = 2*alphas[,K,j] - alphas[,K-1,j],
                            sd = sigmas[,j])
  second_next_alpha <- rnorm(n = nsims,
                             mean = 2*first_next_alpha - alphas[,K,j],
                             sd = sigmas[,j])
  
  alphas_proj[,1,j] <- first_next_alpha
  alphas_proj[,2,j] <- second_next_alpha
  # now project the rest
  for(i in 3:proj_steps){
    alphas_proj[,i,j] <- rnorm(n = nsims,
                               mean = 2*alphas_proj[,i-1,j] - alphas_proj[,i-2,j],
                               sd = sigmas[,j])
  }
}

# now use these to get y's
y_proj <- array(NA, c(nsims, length(proj_years), length(states)))
for(i in 1:length(proj_years)){
  for(j in 1:length(states)){
    all_alphas <- cbind(alphas[,,j], alphas_proj[,,j] )
    this_lambda <- all_alphas %*% as.matrix(Bs_full[length(years)+i, ])
    #y_proj[,i,j] <- rnorm(n = nsims, mean = this_lambda, sd = sigma_ys[,j])
    y_proj[,i,j] <- this_lambda
  }
}

proj_df <- tibble()

for(i in 1:length(proj_years)){
  for(j in 1:length(states)){
    proj_df<-  bind_rows(proj_df, tibble(year = proj_years[i],
                                         state = states[j],
                                         median = exp(quantile(y_proj[,i,j], 0.5)),
                                         lower = exp(quantile(y_proj[,i,j], 0.65)),
                                         upper = exp(quantile(y_proj[,i,j], 0.35))))
  }
}


d %>%
  dplyr::select(state, year, ent_pc) %>%
  mutate(s = rep(1:length(states), each = nrow(y))) %>%
  left_join(yhat %>%   ungroup() %>% dplyr::select(s,year, fit:fit.upper)) %>%
  filter(state %in% c("Texas", "California", "Mississippi", "Ohio")) %>%
  ggplot(aes(year, ent_pc)) +
  geom_point(aes(color = state)) +
  geom_line(aes(year, fit, color = state)) +
  geom_ribbon(aes(ymin = fit.lower, ymax = fit.upper, fill = state), alpha = 0.2) +
  geom_line(data = proj_df %>% filter(year < 2026, state %in% c("Texas", "California", "Mississippi", "Ohio")), aes(year, median, color = state)) +
  geom_ribbon(data = proj_df %>% filter(year < 2026,state %in% c("Texas", "California", "Mississippi", "Ohio")), aes(year, median, ymin = lower, ymax =  upper, fill = state), alpha = 0.2) +
  theme_bw() + ggtitle("Estimated and projected entries per capita \nsecond-order P-splines") + ylab("")
```

