---
title: "Week 4: Gibbs Sampling and Webscraping"
date: today
date-format: "DD/MM/YY"
format: pdf
execute: 
  warning: false
  message: false
---

# Gibbs Sampling

In the first part of this lab we will write a Gibbs Sampler to get posterior estimates of the mean and variance of kid's test scores. Firstly, let's read in the data:

```{r}
library(tidyverse)
kidiq <- read_rds("../data/kidiq.RDS")
```

Set up data, and choose some values for the prior hyperparameters: 

```{r}
y <- kidiq$kid_score
mean_y <- mean(y)
var_y <- var(y)
sd_y <- sd(y)
n <- length(y)

# prior settings
mu0 <- 0
nu0<-5
sigma_mu0 <-  sigma_0 <- 100
nun <- nu0+n
sigma2_mu0 <- sigma_mu0^2
sigma2_0 <- sigma_0^2
```

Do the sampling (fill in the gaps!)

```{r, eval = F}
set.seed(123)
S <-1000
phi <- c(mean_y, 1/var_y)
phi_sp <- matrix(nrow=S,ncol=2) # matrix with dim (nsim, npar)
phi_sp[1,] <- phi
for(s in 2:S) {
  # generate a new mu value from its full conditional

  # generate a new 1/sigma^2 value from its full conditional

  phi_sp[s,] <- phi
}
```

Look at quantiles:
```{r, eval = F}
quantile(phi_sp[,1],c(.025,.5,.975))
quantile(phi_sp[,2],c(.025,.5, .975))
quantile(1/sqrt(phi_sp[,2]),c(.025,.5, .975))
```


Plot the samples:

```{r, eval = F}
par(lwd = 3, cex.axis = 1.5, cex.lab = 2, cex.main = 2, mfrow = c(1,3), mar=c(5,5,1,1))
for (m1 in c(5,15,100)){
  plot( phi_sp[1:m1,],type="l",xlim=range(phi_sp[1:100,1]), ylim=range(phi_sp[1:100,2]),
        lty=1,col="gray",xlab=expression(mu),ylab=expression(1/sigma^2))
  text(  phi_sp[1:m1,1], phi_sp[1:m1,2], c(1:m1), cex = 2)
}
```


## Exercises

- Play around with different initial values
- Plot traceplots

# Webscraping

## Introduction

There's a bunch of different ways to web-scrape, but we'll be exploring using the `rvest` package in R, that helps you to deal with parsing html. 

Why is web scraping useful? If our research involves getting data from a website that isn't already in a easily downloadable form, it improves the reproducibility of our research. Once you get a scraper working, it's less prone to human error than copy-pasting, for example, and much easier for someone else to see what you did. 

## A note on responsibility

Seven principles for web-scraping responsibly:

1. Try to use an API.
2. Check robots.txt. (e.g. https://www.utoronto.ca/robots.txt)
3. Slow down (why not only visit the website once a minute if you can just run your data collection in the background while you're doing other things?).
4. Consider the timing (if it's a retailer then why not set your script to run overnight?).
5. Only scrape once (save the data as you go and monitor where you are up to).
6. Don't republish the data you scraped (cf datasets that create based off it).
7. Take ownership (add contact details to your scripts, don't hide behind VPNs, etc)

##  Extracting the Billboard top 100 songs list


We're going to scrape the current Billboard top 100 songs list https://www.billboard.com/charts/hot-100/. While the data are nicely presented, there's no nice way of downloading the data for each year as a csv or similar form. So let's use `rvest` to extract the data. We'll also load in `janitor` to clean up column names etc later on. 

```{r}
library(tidyverse)
library(rvest)
library(janitor)
```

Firstly, read the html from the webpage:


```{r}
link <- "https://www.billboard.com/charts/hot-100/"
billboard_page <- read_html(link)
```

We want to extract information from the (interactive) table. To find where this was in the html, I made use of the 'Inspect' option when you right-click in Chrome. This was a relatively straightforward way of find the right class for each table item. Once I had that, I pulled all the html nodes of that class, converted to text, removed all "\t" (tabs), converted to a tibble and then separated the values. The final line of code gets rid of all empty columns. 

```{r}
# class of table objects
class <- "o-chart-results-list-row-container"

rough_table <- billboard_page |> 
  html_nodes(xpath = paste0("//div[@class = '",class,"']"))|> # pull out table
  html_text() |> # convert to text
  str_remove_all("\t") |> # remove tabs
  as_tibble() |> # convert to tibble
  separate(value, into = paste0(1:1000), sep = "\n") |> # separate into columns
  select_if(function(x) !(all(is.na(x)) | all(x=="")))  # remove empty columns

```

We're almost there, but some of the values are mis-aligned, because some songs have an additional label of 'new' or a 're-entry'. To tidy this up, I found it easiest to separate these songs out, clean up the table, and then join them back together.

```{r}
# new songs tidying up
rough_new <- rough_table |> 
  filter(`12`=="NEW")|> 
  select_if(function(x) !(all(is.na(x)) | all(x==""))) |> 
  rename(position = `5`,
         song_name = `43`,
         artist = `48`,
         peak_position = `79`,
         weeks_in_chart = `92`
         ) |> 
    mutate(last_week = "-") |> 
  select(position, song_name, artist, last_week, peak_position, weeks_in_chart) |> 
  mutate(label = "new")

# re-entry songs tidying up
rough_re <- rough_table |> 
  filter(`12` == "RE-") |> 
  mutate(`49` = ifelse(`49`=="", `50`, `49`),
         `80` = ifelse(`80`=="", `82`, `80`),
         `93` = ifelse(`93`=="", `95`, `93`))|> 
  rename(position = `5`,
         song_name = `44`,
         artist = `49`,
         peak_position = `80`,
         weeks_in_chart = `93`
         ) |> 
  mutate(last_week = "-") |> 
  select(position, song_name, artist, last_week, peak_position, weeks_in_chart) |> 
  mutate(label = "re-entry")

# everything else tidying up
rough_everything_else <- rough_table |> 
  filter(`12` == "") |> 
  mutate(`44` = ifelse(`44`=="", `45`, `44`),
        `60` = ifelse(`60`=="", `62`, `60`),
         `86` = ifelse(`86`=="", `88`, `86`),
         `73` = ifelse(`73`=="", `75`, `73`))|> 
  rename(position = `5`,
         song_name = `39`,
         artist = `44`,
         last_week = `60`,
         peak_position = `73`,
         weeks_in_chart = `86`
         ) |> 
  select(position, song_name, artist, last_week, peak_position, weeks_in_chart) |> 
  mutate(label = "NA")

# bind them all together and sort
clean_table <- bind_rows(rough_everything_else, 
                         rough_new, rough_re) |> 
  mutate(position = as.numeric(position)) |> 
  arrange(position)

clean_table
```


# Install rstan and brms

We will be using the packages `rstan` and `brms` from next week. Please install these. Here's some instructions:

- https://github.com/paul-buerkner/brms
- https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started

In most cases it will be straightforward and may not need much more than `install.packages()`, but you might run into issues. Every Stan update seems to cause problems for different OS.

To make sure it works, run the following code:

```{r}
library(brms)

x <- rnorm(100)
y <- 1 + 2*x + rnorm(100)
d <- tibble(x = x, y= y)

mod <- brm(y~x, data = d)
summary(mod)

```







